Spatial Reasoning and Statistical Graphics

### 2014-11-05 Paper/Draft1/TestingVisualAptitude.Rnw
latex issues fixed

<<ResultsScaledScores, echo=F, include=T, fig.pos="htb", fig.cap="Test scores for lineups and visuospatial tests. As none of the participants scored at or below zero, we can conclude that there is little evidence of random guessing. We also note the score compression that occurs on the Visual Search test; this indicates that most participants scored extremely high, and thus, participants' scores are not entirely representative of their ability. \\label{fig:Scores}", fig.width=4.5, fig.height=4.5, out.width='\\linewidth', warning=F, message=F>>=

@@

ggplot() + 
  geom_bar(data=qrange, aes(width=.9, x=Test, y=n, fill=Type), alpha=.35, stat="identity", position="identity", colour="red", inherit.aes=F) +
  scale_fill_manual("", values=c(NA, "red")) + 
  geom_boxplot(data=score.summary, aes(x=factor(Test), y=Score), inherit.aes=F, fill="grey") +
  theme_bw()  + 
  ylab("Test Scores") + 
  xlab("") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  theme(plot.margin=unit(c(0,0,0,0),"mm")) + 
  theme(legend.position=c(.68,.125),  legend.box.just = "left", 
        legend.key.height=unit(6, "mm")) 
  #coord_flip()
@
%

@@

Incorporating the lineup task into the principal component analysis, we find the principal components to be fairly similar to the four-component analysis. 
Table~\ref{tab:PCAvariance5} shows the importance of each principal component. \newtext{From the distribution of the variance components, we see that the lineup test  spans an additional dimension within the space of the four established tests.} From the rotation matrix (see Table~\ref{tab:PCArotation5}) we see that

<<pcasummary5,echo=F, results='asis'>>=
print(xtable(summary(pca), caption="Importance of principal components, analyzing all five tests.\\label{tab:PCAvariance5}",digits=2), sanitize.text.function=sanitize, sanitize.rownames.function=sanitize, sanitize.colnames.function=sanitize, size="footnotesize", table.placement="htb", caption.placement="top")
@

<<pcarotation5,echo=F, results='asis'>>=
print(xtable(pca$rotation, caption="PCA Rotation matrix for all five tests. The first principal component is essentially an average of all five tests. 
             \\label{tab:PCArotation5}",digits=2), sanitize.text.function=sanitize, sanitize.rownames.function=sanitize, sanitize.colnames.function=sanitize, table.placement="htb", caption.placement="top")
@

the first principal component, PC1, is again essentially an average across all tests and accounts for \Sexpr{round(summary(pca)$importance[3,1]*100,1)}\% of the variance in the data. 
Biplots of the remaining components are provided in Appendix \ref{app:pca5}.

@@

# use AIC in backward selection
library(MASS)
m1 <- stepAIC(m0, direction="both")
@
\newtext{We fit a linear model of lineup scores in all of the demographic variables and the test scores from the visuo-spatial tests. Then we use a step-wise backwards selection based on AIC (Akaike Information Criterion - do we need a reference?) to select the best model. The result is shown in table~\ref{tab:m1}. Only two covariates stay in the model: PC1 and MATH, reflecting two dimensions of what affects lineup scores. We can think of PC1 as an innate visual ability, while the MATH score is a matter of both ability and training. 
}
<<modelresults, echo=FALSE, results='asis'>>=
library(xtable)
xtable(summary(m1), caption="Estimates and significances of a linear model of lineup scores.", label="tab:m1")
@
\subsection{Lineup Types}
\comment{
results need to go here
}

@@

This scaling gets slightly more complicated for scores which have two sub-groups, as with the figure classification test, which separately sumarizes male and female participants' scores. 
To get a single unified score with standard deviation, we completed the following calculations: 
\begin{align}
\mu_{\text{all}} &= (N_F\mu_{F} + N_M\mu_{M})/(N_F + N_M)\\
\sigma_{\text{all}} &= \sqrt{\left(N_F\sigma_F^2 + N_M\sigma_M^2\right)/(N_F + N_M)},
\end{align}
\newtext{where $\mu_F$ and $\mu_M$ are the mean of female and male scores respectively, $N_F$ and $N_M$ are the number of participants in each group, and $\sigma_F^2$ and $\sigma_M^2$ are the variance of each group.
Substituting in the provided numbers, we get}
\begin{align*}
\mu_{\text{all}} &= \left(323\!*\!114.9\! +\! 294\!*\!120.0\right)/(323\!+\!294) \\
& = \Sexpr{round((120.0*294 + 114.9*323)/617/2, 1)}\\
\sigma_{\text{all}} &= \sqrt{\left(323(27.8)^2 \!+\! 294(30)^2\right)/(323\!+\!294)} \\
& = \Sexpr{round(sqrt((30^2*294 + 27.8^2*323)/617)/2, 1)}.
\end{align*}

>>>>

\begin{figure}[ht]
<<ResultsScaledScores, echo=F, include=T, fig.width=4.5, fig.height=4.5, out.width='\\linewidth', warning=F, message=F>>=

@@

ggplot() + 
  geom_bar(data=qrange, aes(width=.9, x=Test, y=n, fill=Type), alpha=.35, stat="identity", position="identity", colour="red", inherit.aes=F) +
  scale_fill_manual("", values=c(NA, "red")) + 
  geom_boxplot(data=score.summary, aes(x=factor(Test), y=Score), inherit.aes=F, fill="grey") +
  theme_bw()  + 
  ylab("Test Scores") + 
  xlab("") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  theme(plot.margin=unit(c(0,0,0,0),"mm")) + 
  theme(legend.position=c(.68,.125),  legend.box.just = "left", 
        legend.key.height=unit(6, "mm")) 
  #coord_flip()
@
\caption{Test scores for lineups and visuospatial tests. As none of the participants scored at or below zero, we can conclude that there is little evidence of random guessing. We also note the score compression that occurs on the Visual Search test; this indicates that most participants scored extremely high, and thus, participants' scores are not entirely representative of their ability. \label{fig:Scores}}
\end{figure}
%

@@

Incorporating the lineup task into the principal component analysis, we find the principal components to be fairly similar to the four-component analysis. 
Table~\ref{tab:PCAvariance5} shows the importance of each principal component. From the distribution of the variance components, we see that the lineup test  spans an additional dimension within the space of the four established tests. 
<<pcasummary5,echo=F, results='asis'>>=
print(xtable(summary(pca), caption="Importance of principal components, analyzing all five tests.\\label{tab:PCAvariance5}",digits=2), sanitize.text.function=sanitize, sanitize.rownames.function=sanitize, sanitize.colnames.function=sanitize, size="footnotesize", table.placement="ht", caption.placement="top")
@

<<pcarotation5,echo=F, results='asis'>>=
print(xtable(pca$rotation, caption="PCA Rotation matrix for all five tests. The first principal component is essentially an average of all five tests. 
             \\label{tab:PCArotation5}",digits=2), sanitize.text.function=sanitize, sanitize.rownames.function=sanitize, sanitize.colnames.function=sanitize, table.placement="htb", caption.placement="top")
@
From the rotation matrix (see Table~\ref{tab:PCArotation5}) we see that  
the first principal component, PC1, is again essentially an average across all tests and accounts for \Sexpr{round(summary(pca)$importance[3,1]*100,1)}\% of the variance in the data. 
Biplots of the remaining components are provided in Appendix \ref{app:pca5}.

@@

# use AIC in backward selection
library(MASS)
m1 <- stepAIC(m0, direction="both")
@
\newtext{We fit a linear model of lineup scores in all of the demographic variables and the test scores from the visuo-spatial tests. Then we use a step-wise backwards selection based on AIC (Akaike Information Criterion - do we need a reference?) to select the best model. The result is shown in table~\ref{tab:m1}. Only two covariates stay in the model: PC1 and MATH, reflecting two dimensions of what affects lineup scores. We can think of PC1 as an innate visual ability, while the MATH score is a matter of both ability and training. 
}
<<modelresults, echo=FALSE, results='asis'>>=
library(xtable)
print(xtable(summary(m1), caption="Estimates and significances of a linear model of lineup scores.", label="tab:m1"), caption.placement="top")
@
\subsection{Lineup Types}
\comment{
results need to go here
}

@@

This scaling gets slightly more complicated for scores which have two sub-groups, as with the figure classification test, which separately sumarizes male and female participants' scores. 
To get a single unified score with standard deviation, we completed the following calculations: 
\begin{align}
\mu_{\text{all}} &= (N_F\mu_{F} + N_M\mu_{M})/(N_F + N_M)\\
\sigma_{\text{all}} &= \sqrt{\left(N_F\sigma_F^2 + N_M\sigma_M^2\right)/(N_F + N_M)},
\end{align}
\newtext{where $\mu_F$ and $\mu_M$ are the mean of female and male scores respectively, $N_F$ and $N_M$ are the number of participants in each group, and $\sigma_F^2$ and $\sigma_M^2$ are the variance of each group.
Substituting in the provided numbers, we get}
\begin{align*}
\mu_{\text{all}} &= \left(323\!\cdot\!114.9\! +\! 294\!\cdot\!120.0\right)/(323\!+\!294) \\
& = \Sexpr{round((120.0*294 + 114.9*323)/617/2, 1)}\\
\sigma_{\text{all}} &= \sqrt{\left(323(27.8)^2 \!+\! 294(30)^2\right)/(323\!+\!294)} \\
& = \Sexpr{round(sqrt((30^2*294 + 27.8^2*323)/617)/2, 1)}.
\end{align*}
